Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-30 manuscript/manuscript.Rmd
minor revisions in the write-up.

Because all of these perspectives have some empirical support, the current debate is not about whether the processes they posit exist at all, but about their relative contribution to explaining the cultural differences we see in the world. Nevertheless, researchers have struggled to reach a consensus on the importance of cultural change during adulthood. Over the span of a few years, most apparent changes in adults' survey responses appear to be transitory, with little evidence of persistent change [@kiley2020; @vaisey2021]. This suggests that change during adulthood may not be a major factor in explaining contemporary cultural divides. By contrast, when we consider a longer time horizon, there is evidence that adults make at least some persistent changes [@lersch2023].

@@

Other studies are compatible with these findings as well. For example, Brocic and Miles [-@brocic2021] estimate that completing graduate degrees in humanities, arts, and social sciences shift peoples' moral relativism only about .2 standard deviations on average compared to people with no college degree, and this is the largest effect they identify. And studies of aggregate change show that, even on items where cohort effects explain more variance than period effects, there is always evidence that some people change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.

@@

To this point, the theoretical debate has been about whether people *ever* change. As we argued above, this is rarely the right question since the answer is almost always "yes." Over sufficient time and in a large enough sample, researchers will likely observe some evidence for durable individual change. A lack of evidence for change may be due to a poor survey question, a low resolution in response options, or that the question simply was not asked for long enough or to enough people. Conversely, finding evidence for some change tells us only that it was "not zero." From that alone, we learn little about how much intrapersonal change has happened in a population.

@@

Although we did not state expectations for how question content would relate to $\omega$, some of the associations between question structure, panel duration, and panels themselves might be driven by differences in the topics addressed by each panel. To address this, we followed Hout et al. [-@hout2016] and Lersch [-@lersch2023] in coding each question as falling into one of nine different content categories and included these as indicator variables in the meta analysis. Supplemental Materials B presents the estimates for these content indicators. They show some associations with $\omega$, with questions about subjective SES; social life, social cohesion, and trust; environment and climate; and health and morale showing larger $\omega$ values than questions about religion and spirituality; politics, government and the economy; and gender and family.

The most notable result in the meta analysis is the negative coefficient attached to the duration of years covered by the question. The longer a question is observed, the less is the value of $\omega$. Theories of personal cultural change that link changes in personal culture to social experiences, including the LCAM, suggest that the longer we observe respondents, the more likely people are to undergo potentially transformative experiences and therefore the more variance would be attributable to intrapersonal change. Finding a negative coefficient here seems to challenge that assumption.

As a further test of this finding, we compared the $\omega$ values when using the full duration of a panel compared to when we dropped the final wave and therefore reduced the total duration of observation for the question. If the coefficient reflects a true negative effect of duration on $\omega$, we should see that same effect within questions. Results from this analysis are presented in Appendix B and contradict the coefficient from the regression model; with a few exceptions, we found that the longer we observe the same question, the higher $\omega$ is. We interpret this combination of findings as suggesting that the kinds of questions asked for a longer time period tend to demonstrate less intrapersonal change than questions asked over shorter periods, rather than a true function of time. This supports the conclusion that the GSS shows less intrapersonal change because of its shorter duration.

@@

```{=latex}
\end{center}
\footnotesize{\textit{Notes:} $\omega$ values across college graduates and college non-graduates on 8 political culture items from the General Social Survey (2006-2014).}
\end{figure}
```
On all eight questions presented in Figure 4, $\omega$ values are smaller for people with a college degree, meaning intrapersonal change is less common among them. There are also large differences in $\omega$ values across questions for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than one percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

@@

Figure A2 presents the alternative strategy we used to understand the effects of duration on intraindividual change. First, we restricted the sample such that we have at least 4 waves of observations for all participants. This reduced the number of items to 249. We then fitted the Life Course Adaption models to this sample. Second, we removed the last wave from all the observations, and refit the same models once more. In the final step, we calculated the $\omega$ values for each item. The Figure C1 shows a scatterplot for these two sets of observations.

>>>>

Because all of these perspectives have some empirical support, the theoretical debate is not about whether the processes they posit exist at all, but about their relative contribution to explaining the cultural differences we see in the world. Nevertheless, researchers have struggled to reach a consensus on the importance of cultural change during adulthood. Over the span of a few years, most apparent changes in adults' survey responses appear to be transitory, with little evidence of persistent change [@kiley2020; @vaisey2021]. This suggests that change during adulthood may not be a major factor in explaining contemporary cultural divides. By contrast, when we consider a longer time horizon, there is evidence that adults make at least some persistent changes [@lersch2023].

@@

Other studies are compatible with these findings as well. For example, Brocic and Miles [-@brocic2021] estimate that completing graduate degrees in humanities, arts, and social sciences shift peoples' moral relativism only about 0.2 standard deviations on average compared to people with no college degree, and this is the largest effect they identify. And studies of aggregate change show that, even on items where cohort effects explain more variance than period effects, there is always evidence that some people change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.

@@

To this point, the debate has been about whether people *ever* change. As we argued above, this is rarely the right question since the answer is almost always "yes." Over sufficient time and in a large enough sample, researchers will likely observe some evidence for durable individual change. A lack of evidence for change may be due to a poor survey question, a low resolution in response options, or that the question simply was not asked for long enough or to enough people. Conversely, finding evidence for some change tells us only that it was "not zero." From that alone, we learn little about how much intrapersonal change has happened in a population.

@@

Although we did not state expectations for how question content would relate to $\omega$, some of the associations between question structure, panel duration, and panels themselves might be driven by differences in the topics addressed by each panel. To address this, we followed Hout et al. [-@hout2016] and Lersch [-@lersch2023] in coding each question as falling into one of nine different topical domains and included these as indicator variables in the meta analysis. Supplemental Materials B presents the estimates for these topic indicators. They show some associations with $\omega$, with questions about subjective SES; social life, social cohesion, and trust; environment and climate; and health and morale showing larger $\omega$ values than questions about religion and spirituality; politics, government and the economy; and gender and family.

The most notable result in the meta analysis is the negative coefficient attached to the duration of years covered by the question on average across participants. The longer a question is observed, the less is the value of $\omega$. Theories of personal cultural change that link changes in personal culture to social experiences, including the LCAM, suggest that the longer we observe respondents, the more likely people are to undergo potentially transformative experiences and therefore the more variance would be attributable to intrapersonal change. Finding a negative coefficient here seems to challenge that assumption.

As a further test of this finding, we compared the $\omega$ values when using the full duration of a panel compared to when we dropped the final wave for each participant and therefore reduced the total duration of observation for the question. If the coefficient reflects a true negative effect of duration on $\omega$, we should see that same effect within questions. Results from this analysis are presented in Appendix B and contradict the coefficient from the regression model; with a few exceptions, we found that the longer we observe the same question, the higher $\omega$ is. We interpret this combination of findings as suggesting that the kinds of questions asked for a longer time period tend to demonstrate less intrapersonal change than questions asked over shorter periods, rather than a true function of time. This supports the conclusion that the GSS shows less intrapersonal change because of its shorter duration.

@@

```{=latex}
\end{center}
\footnotesize{\textit{Notes:} $\omega$ values across college graduates and college non-graduates on 8 political culture items from the General Social Survey (2006-2014).}
\end{figure}
```
On all eight questions presented in Figure 4, $\omega$ values are smaller for people with a college degree, meaning intrapersonal change is less common among them. There are also large differences in $\omega$ values across questions for both groups. For example, on the question of whether the government should try to solve more problems or leave those problems to be solved by private businesses ("government do more or less"), less than 1 percent of the systematic variance is attributable to intrapersonal change for both groups. In other words, while people might vacillate on this question at random (37 percent of variance is residual for this question), there is functionally no evidence that people make systematic changes of opinion on this issue during the GSS panel.

@@

Figure A2 presents the alternative strategy we used to understand the effects of duration on intraindividual change. First, we removed the last wave from all the observations and fitted the Life Course Adaption models. This effectively reduced the number of items to 249. We then refitted the model for all participants in this sample using the unrestricted data. In the final step, we calculated the $\omega$ values for each item. The Figure A2 shows a scatterplot for these two sets of observations.

### 2023-09-30 manuscript/manuscript.tex
minor revisions in the write-up.

Because all of these perspectives have some empirical support, the
current debate is not about whether the processes they posit exist at
all, but about their relative contribution to explaining the cultural
differences we see in the world. Nevertheless, researchers have
struggled to reach a consensus on the importance of cultural change
during adulthood. Over the span of a few years, most apparent changes in
adults' survey responses appear to be transitory, with little evidence
of persistent change (Kiley and Vaisey 2020; Vaisey and Kiley 2021).
This suggests that change during adulthood may not be a major factor in
explaining contemporary cultural divides. By contrast, when we consider
a longer time horizon, there is evidence that adults make at least some
persistent changes (Lersch 2023).

@@

Other studies are compatible with these findings as well. For example,
Brocic and Miles (2021) estimate that completing graduate degrees in
humanities, arts, and social sciences shift peoples' moral relativism
only about .2 standard deviations on average compared to people with no
college degree, and this is the largest effect they identify. And
studies of aggregate change show that, even on items where cohort
effects explain more variance than period effects, there is always
evidence that some people change over time (Vaisey and Lizardo 2016). In
other words, despite different interpretations, the results of previous
work are in many ways highly consonant.

@@

To this point, the theoretical debate has been about whether people
\emph{ever} change. As we argued above, this is rarely the right
question since the answer is almost always ``yes.'' Over sufficient time
and in a large enough sample, researchers will likely observe some
evidence for durable individual change. A lack of evidence for change
may be due to a poor survey question, a low resolution in response
options, or that the question simply was not asked for long enough or to
enough people. Conversely, finding evidence for some change tells us
only that it was ``not zero.'' From that alone, we learn little about
how much intrapersonal change has happened in a population.

@@

Although we did not state expectations for how question content would
relate to \(\omega\), some of the associations between question
structure, panel duration, and panels themselves might be driven by
differences in the topics addressed by each panel. To address this, we
followed Hout et al. (2016) and Lersch (2023) in coding each question as
falling into one of nine different content categories and included these
as indicator variables in the meta analysis. Supplemental Materials B
presents the estimates for these content indicators. They show some
associations with \(\omega\), with questions about subjective SES;
social life, social cohesion, and trust; environment and climate; and
health and morale showing larger \(\omega\) values than questions about
religion and spirituality; politics, government and the economy; and
gender and family.

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question. The longer a
question is observed, the less is the value of \(\omega\). Theories of
personal cultural change that link changes in personal culture to social
experiences, including the LCAM, suggest that the longer we observe
respondents, the more likely people are to undergo potentially
transformative experiences and therefore the more variance would be
attributable to intrapersonal change. Finding a negative coefficient
here seems to challenge that assumption.

As a further test of this finding, we compared the \(\omega\) values
when using the full duration of a panel compared to when we dropped the
final wave and therefore reduced the total duration of observation for
the question. If the coefficient reflects a true negative effect of
duration on \(\omega\), we should see that same effect within questions.
Results from this analysis are presented in Appendix B and contradict
the coefficient from the regression model; with a few exceptions, we
found that the longer we observe the same question, the higher
\(\omega\) is. We interpret this combination of findings as suggesting
that the kinds of questions asked for a longer time period tend to
demonstrate less intrapersonal change than questions asked over shorter
periods, rather than a true function of time. This supports the
conclusion that the GSS shows less intrapersonal change because of its
shorter duration.

@@

On all eight questions presented in Figure 4, \(\omega\) values are
smaller for people with a college degree, meaning intrapersonal change
is less common among them. There are also large differences in
\(\omega\) values across questions for both groups. For example, on the
question of whether the government should try to solve more problems or
leave those problems to be solved by private businesses (``government do
more or less''), less than one percent of the systematic variance is
attributable to intrapersonal change for both groups. In other words,
while people might vacillate on this question at random (37 percent of
variance is residual for this question), there is functionally no
evidence that people make systematic changes of opinion on this issue
during the GSS panel.

@@

Figure A2 presents the alternative strategy we used to understand the
effects of duration on intraindividual change. First, we restricted the
sample such that we have at least 4 waves of observations for all
participants. This reduced the number of items to 249. We then fitted
the Life Course Adaption models to this sample. Second, we removed the
last wave from all the observations, and refit the same models once
more. In the final step, we calculated the \(\omega\) values for each
item. The Figure C1 shows a scatterplot for these two sets of
observations.

>>>>

Because all of these perspectives have some empirical support, the
theoretical debate is not about whether the processes they posit exist
at all, but about their relative contribution to explaining the cultural
differences we see in the world. Nevertheless, researchers have
struggled to reach a consensus on the importance of cultural change
during adulthood. Over the span of a few years, most apparent changes in
adults' survey responses appear to be transitory, with little evidence
of persistent change (Kiley and Vaisey 2020; Vaisey and Kiley 2021).
This suggests that change during adulthood may not be a major factor in
explaining contemporary cultural divides. By contrast, when we consider
a longer time horizon, there is evidence that adults make at least some
persistent changes (Lersch 2023).

@@

Other studies are compatible with these findings as well. For example,
Brocic and Miles (2021) estimate that completing graduate degrees in
humanities, arts, and social sciences shift peoples' moral relativism
only about 0.2 standard deviations on average compared to people with no
college degree, and this is the largest effect they identify. And
studies of aggregate change show that, even on items where cohort
effects explain more variance than period effects, there is always
evidence that some people change over time (Vaisey and Lizardo 2016). In
other words, despite different interpretations, the results of previous
work are in many ways highly consonant.

@@

To this point, the debate has been about whether people \emph{ever}
change. As we argued above, this is rarely the right question since the
answer is almost always ``yes.'' Over sufficient time and in a large
enough sample, researchers will likely observe some evidence for durable
individual change. A lack of evidence for change may be due to a poor
survey question, a low resolution in response options, or that the
question simply was not asked for long enough or to enough people.
Conversely, finding evidence for some change tells us only that it was
``not zero.'' From that alone, we learn little about how much
intrapersonal change has happened in a population.

@@

Although we did not state expectations for how question content would
relate to \(\omega\), some of the associations between question
structure, panel duration, and panels themselves might be driven by
differences in the topics addressed by each panel. To address this, we
followed Hout et al. (2016) and Lersch (2023) in coding each question as
falling into one of nine different topical domains and included these as
indicator variables in the meta analysis. Supplemental Materials B
presents the estimates for these topic indicators. They show some
associations with \(\omega\), with questions about subjective SES;
social life, social cohesion, and trust; environment and climate; and
health and morale showing larger \(\omega\) values than questions about
religion and spirituality; politics, government and the economy; and
gender and family.

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question on average
across participants. The longer a question is observed, the less is the
value of \(\omega\). Theories of personal cultural change that link
changes in personal culture to social experiences, including the LCAM,
suggest that the longer we observe respondents, the more likely people
are to undergo potentially transformative experiences and therefore the
more variance would be attributable to intrapersonal change. Finding a
negative coefficient here seems to challenge that assumption.

As a further test of this finding, we compared the \(\omega\) values
when using the full duration of a panel compared to when we dropped the
final wave for each participant and therefore reduced the total duration
of observation for the question. If the coefficient reflects a true
negative effect of duration on \(\omega\), we should see that same
effect within questions. Results from this analysis are presented in
Appendix B and contradict the coefficient from the regression model;
with a few exceptions, we found that the longer we observe the same
question, the higher \(\omega\) is. We interpret this combination of
findings as suggesting that the kinds of questions asked for a longer
time period tend to demonstrate less intrapersonal change than questions
asked over shorter periods, rather than a true function of time. This
supports the conclusion that the GSS shows less intrapersonal change
because of its shorter duration.

@@

On all eight questions presented in Figure 4, \(\omega\) values are
smaller for people with a college degree, meaning intrapersonal change
is less common among them. There are also large differences in
\(\omega\) values across questions for both groups. For example, on the
question of whether the government should try to solve more problems or
leave those problems to be solved by private businesses (``government do
more or less''), less than 1 percent of the systematic variance is
attributable to intrapersonal change for both groups. In other words,
while people might vacillate on this question at random (37 percent of
variance is residual for this question), there is functionally no
evidence that people make systematic changes of opinion on this issue
during the GSS panel.

@@

Figure A2 presents the alternative strategy we used to understand the
effects of duration on intraindividual change. First, we removed the
last wave from all the observations and fitted the Life Course Adaption
models. This effectively reduced the number of items to 249. We then
refitted the model for all participants in this sample using the
unrestricted data. In the final step, we calculated the \(\omega\)
values for each item. The Figure A2 shows a scatterplot for these two
sets of observations.
